---
title: "Working with HCES Nesstar files using nesstarR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with HCES Nesstar files using nesstarR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

# pkgdown sets opts_knit$root.dir to the package root before knitting.
# When rendering locally from vignettes/, ".." resolves to the package root.
pkg_root <- {
  rd <- knitr::opts_knit$get("root.dir")
  if (!is.null(rd) && nzchar(rd)) normalizePath(rd) else normalizePath("..")
}
nesstar_path <- file.path(
  pkg_root, "data", "hces",
  "DDI-IND-NSSO-66-SCHEDULE-1.0T2",
  "Nss66_1.0-type2_new format-1",
  "survey0", "data",
  "nss66_consumer_expenditure_type_2.Nesstar"
)
```

NSS rounds 64, 66, and 68 come as `.Nesstar` binaries packed inside `.rar` archives.
This vignette covers downloading one round with mospiR, extracting the archive, and
reading the data with [nesstarR](https://github.com/saketkc/nesstarR).

## Prerequisites

```{r install, eval=FALSE}
remotes::install_github("saketkc/mospiR")
remotes::install_github("saketkc/nesstarR")
```

```{r libs}
library(mospiR)
library(nesstarR)
```

## Download from the portal

```{r download, eval=FALSE}
api_key <- Sys.getenv("MOSPI_KEY")

download_dataset(
  "DDI-IND-NSSO-66-SCHEDULE-1.0T2",
  file.path("data", "hces", "DDI-IND-NSSO-66-SCHEDULE-1.0T2"),
  api_key
)
```

The download is a `.rar` archive, about 54 MB for NSS 66 T2.

## Extract the archive

`unar` (macOS/Linux) and `unrar` (Windows) both work. Either must be on your `PATH`.

```{r extract, eval=FALSE}
rar_file <- file.path(
  "data", "hces", "DDI-IND-NSSO-66-SCHEDULE-1.0T2",
  "Nss66_1.0-type2_new format.rar"
)
system2("unar", c("-o", dirname(rar_file), shQuote(rar_file)))
```

Extraction produces a folder with the `.Nesstar` binary, `ddi.xml` (variable metadata),
and supporting documents.

## Parse the Nesstar file

`nesstar_parse()` reads the binary header without loading any row data. Data loads
only when you call `nesstar_read_dataset()` on a specific dataset number.

```{r parse}
nb <- nesstar_parse(nesstar_path)
nb
```

## Dataset structure

A single `.Nesstar` file holds multiple datasets, one per schedule block.

```{r datasets}
nesstar_datasets(nb)
```

Dataset 21 is the food block: 4.8 million item-level rows across roughly 100,000 households.

## Variable listing

```{r variables}
vars <- nesstar_variables(nb, dataset_number = 21)
vars[, c("name", "variable_id", "width_value")]
```

Key columns in the food block:

| Column | Meaning |
|--------|---------|
| `HH_ID` | Household identifier |
| `State` | State code (2-digit) |
| `State_District` | District code (4-digit: state × 100 + district) |
| `Item_code` | Food item code (NSS 66 coding) |
| `Total_Value` | Household monthly expenditure (Rs) |
| `Multiplier` | Survey weight |

## Read a dataset

```{r read}
food <- nesstar_read_dataset(nb, dataset_number = 21)
cat("Rows:", nrow(food), "| Columns:", ncol(food), "\n")
head(food[, c("HH_ID", "State", "State_District",
              "Item_code", "Total_Value", "Multiplier")])
```

## Quick check: cereal spending by sector

Weighted mean monthly expenditure on cereals (item codes 101-128), rural vs. urban:

```{r explore}
cereals    <- food[food$Item_code >= 101 & food$Item_code <= 128, ]
hh_cereals <- aggregate(Total_Value ~ HH_ID + Sector + Multiplier,
                        data = cereals, FUN = sum)
rural <- hh_cereals[hh_cereals$Sector == 1, ]
urban <- hh_cereals[hh_cereals$Sector == 2, ]

cat(sprintf(
  "Weighted mean cereal expenditure (Rs/month):\n  Rural: %.0f\n  Urban: %.0f\n",
  weighted.mean(rural$Total_Value, rural$Multiplier, na.rm = TRUE),
  weighted.mean(urban$Total_Value, urban$Multiplier, na.rm = TRUE)
))
```

## Export to CSV

`nesstar_export()` writes one CSV per dataset to `output_dir`.

```{r export}
output_dir <- file.path(tempdir(), "nss66t2")
nesstar_export(nb, output_dir = output_dir, compress = FALSE)
list.files(output_dir)
```

Pass `compress = TRUE` for `.csv.gz` output.

## Round reference

| Round | idno | Period | Format |
|-------|------|--------|--------|
| NSS 57th | `DDI-IND-MOSPI-NSSO-57Rnd-Sch1.0-2001` | 2001 | CSV zip |
| NSS 58th | `DDI-IND-MOSPI-NSSO-58Rnd-Sch1.0-2002` | 2002 | CSV zip |
| NSS 59th | `DDI-IND-MOSPI-NSSO-59Rnd-Sch1.0-2003` | 2003 | CSV zip |
| NSS 60th | `DDI-IND-MOSPI-NSSO-60Rnd-Sch1-Jan-June2004` | 2004 | CSV zip |
| NSS 61st | `DDI-IND-MOSPI-NSSO-61Rnd-Sch1-July2004-June2005` | 2004-05 | CSV zip |
| NSS 62nd | `DDI-IND-MOSPI-NSSO-62Rnd-Sch1.0-2005-06` | 2005-06 | CSV zip |
| NSS 63rd | `DDI-IND-MOSPI-NSSO-63Rnd-Sch1.0-2006-07` | 2006-07 | CSV zip |
| NSS 64th | `IND-NSSO-HCES-2007-v1` | 2007-08 | Nesstar |
| NSS 66th T1 | `DDI-IND-NSSO-66-SCHEDULE-1.0T1` | 2009-10 | Nesstar |
| NSS 66th T2 | `DDI-IND-NSSO-66-SCHEDULE-1.0T2` | 2009-10 | Nesstar |
| NSS 68th T1 | `DDI-IND-MOSPI-NSSO-68Rnd-Sch1.0-July2011-June2012` | 2011-12 | Nesstar |
| NSS 68th T2 | `DDI-IND-MOSPI-NSSO-68Rnd-Sch2.0-July2011-June2012` | 2011-12 | Nesstar |

Rounds 57-63 unzip to CSVs. Rounds 64, 66, and 68 need nesstarR.
