---
title: "Getting Started with mospiR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mospiR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = nchar(Sys.getenv("MOSPI_KEY")) > 0
)
api_key <- Sys.getenv("MOSPI_KEY")
```

mospiR downloads survey microdata from the [MoSPI Microdata Portal](https://microdata.gov.in),
India's NSO repository. It wraps the NADA API and manages authentication, pagination, and retries.

## Installation

```{r install, eval=FALSE}
remotes::install_github("saketkc/mospiR")
```

## API key

Register at <https://microdata.gov.in> and copy your key from **My Profile → API Key**.
Add it to `~/.Renviron`:

```
MOSPI_KEY=your_key_here
```

Reload with `readRenviron("~/.Renviron")` or restart R.

```{r key-env}
library(mospiR)
api_key <- Sys.getenv("MOSPI_KEY")
```

## Listing datasets

`list_datasets()` fetches all pages and returns one data frame. Pass `query` to filter by
title (client-side, case-insensitive).

```{r list-all}
datasets <- list_datasets(api_key)
cat("Total datasets:", nrow(datasets), "\n")
head(datasets[, c("idno", "title")], 6)
```

```{r search}
plfs <- list_datasets(api_key, query = "labour force")
plfs[, c("idno", "title")]
```

## Listing files in a dataset

```{r list-files}
files <- list_files("DDI-IND-NSO-ASI-2019-20", api_key)
files[, c("name", "size")]
```

## Downloading files

`download_file()` writes one file to `folder_path` and returns the local path invisibly.

```{r download-file, eval=FALSE}
path <- download_file(
  dataset_id  = "DDI-IND-NSO-ASI-2019-20",
  file_name   = "ASI_DATA_2019_20_CSV.zip",
  folder_path = "data/asi-2019-20",
  api_key     = api_key
)
```

`download_dataset()` downloads every file in the dataset:

```{r download-dataset, eval=FALSE}
paths <- download_dataset(
  dataset_id  = "DDI-IND-NSO-ASI-2019-20",
  folder_path = "data/asi-2019-20",
  api_key     = api_key
)
length(paths)
```

If the file name doesn't match, `download_file()` reports what's available:

```{r download-missing}
result <- download_file("DDI-IND-NSO-ASI-2019-20", "nonexistent.csv",
                        tempdir(), api_key)
```

## HCES rounds

The portal has all rounds of the Household Consumer Expenditure Survey. Filter by title,
then download in a loop:

```{r hces-find}
hces_all <- list_datasets(api_key, query = "consumer expenditure")
cat("Matching datasets:", nrow(hces_all), "\n")
hces_all[, c("idno", "title")]
```

```{r hces-download, eval=FALSE}
for (id in hces_all$idno) {
  download_dataset(id, file.path("data", "hces", id), api_key)
}
```
